Unknown

### 2025-06-20 ms/ms.qmd

@@

<!-- Paragraph 1 -->
Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells. They are essential for growing tall on land because they allow plants to access CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments (citations). Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@boer__2016; harrison__2020; more citations]. Understanding how stomata respond to environmental change over daily, seasonal, and evolutionary time is important for understanding adaptation (cites), predicting paleoclimate from fossil cuticles (cites), and improving crops (cites). Stomatal function contributes to global carbon and water cycles and therefore predicting future climate (cites).

@@

<!-- Paragraph 3 -->
<!-- should i add analysis of herbarium data here? -->
The surprising fact is that despite the potential photosynthetic benefit, most leaves are not amphistomatous. Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake); among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.). Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface. The covariation between stomatal density ratio and light habitat is both qualitative and quantitative. A higher proportion of sun leaves are amphistomatous (salisbury) and the proportion of stomata on the upper, adaxial surface increases with light (muir 2018, bucher?). [end more strongly with statement that this is an important problem to be solved??]

## Outtro - WORKING HERE

It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat. Reaction norms to light intensity among wild tomatoes indicate that accessions from low light habitats express low \gmaxratio{} when grown in shade, whereas accessions from high light habitats have constitutively higher \gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is driven by greater plasticty in leaf structure in accessions from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis. As a consequence, sun leaves benefit more from amphistomy. This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.

## Middle

Leaves with greater stomatal density ratio are more in open, sunny habitats because they deliver the greatest benefit in those circumstances. An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small \[cites\]. We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo). Why would AA be greater in sun than shade? There are three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive'.

Acclimatory hypothesis: Leaves acclimated to high light intensity typically increase total leaf stomatal conductance (increased CO\$\_2\$ supply) and upregulate Rubisco activation <!--# I don't love the wording here. Is it just Rubisco activation, or are there are other things? --> (increased CO\$*2\$* demand). A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information). If the acclimatory hypothesis is correct, we predict that AAhigh \> AAlow for all species regardless of native habitat or growth environment. Plants adapted to sunny, open habitats will evolve greater stomatal density ratio to take advantage of regular exposure to high light intensity.

Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions (cite). Plastic responses are likely adaptations to optimize photosynthesis at different light intensities (cites), but changes in leaf anatomy and biochemistry could modulate AA as a byproduct. Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower g_ias; leaves with increased total stomatal density and Rubisco concentration have greater potential CO2 supply and demand. Under the plastic hypothesis, we predict that AAsun \> AAshade for all species and light intensities. AAsun and gsmax,sun should be positively associated with native light habitat. \[transition\] We assume that genotypes adapted to sunny, open habitats will express a phenotype best adapted to that environment when leaves develop under high light intensity; genotypes adapted to shaded closed habitats may be plastic, but suboptimal for light intensities they do not regularly experience in nature.

\[conceptual figure could also show differential benefit of amphi leaves in sun/shade\]

to plants growing there. in those It has been hypothesized that amphistomy increases photosynthesis more in sunny places. Three nonmutually exclusive adaptive explanations for why amphi leaves:

Conceptual figure explaining each hypothesis

-   acclimation: greater demand, higher gs increase AA
-   developmental plasticity: light-induced changes in leaf anatomy modulate AA
-   constitutive: genetic differences in leaves adapted to different light habitats

@@

>>>>

What's surprising: role of developmental plasticity in explaining comparative results; diffusional limitations through palisade ias is strong and needs to be considered with light gradients in understanding leaf strucuture- funciton

@@

<!-- Paragraph 1 -->
Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells. They are essential for growing tall on land because they allow plants to access CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments (citations). Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@boer__2016; harrison__2020; more citations]. Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation (cites), predicting paleoclimate from fossil cuticles (cites), and improving crops (cites). Stomatal function contributes to global carbon and water cycles and therefore predicting future climate (cites).

@@

<!-- Paragraph 3 -->
The paradoxical fact is that, despite the potential photosynthetic benefit, most leaves are not amphistomatous. Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake). But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.). Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface. Even resupinate leaves develop stomata on the lower, albeit adaxial surface (Lesheyde?), suggesting that leaf orientation (lower vs. upper) rather than leaf polarity (abaxial vs. adaxial) is causal. The covariation between stomatal density ratio and light habitat is both qualitative and quantitative. A higher proportion of sun leaves are amphistomatous (salisbury) and the proportion of stomata on the upper, adaxial surface increases with light (muir 2018, bucher?). Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.

<!-- Paragraph 4+, hypotheses and predictions -->
The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances. An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small (smith paper, muir 2019). We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo) (Parkhurst, Triplett). Why would AA be greater in sun than shade? We consider three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive'.

Acclimatory hypothesis: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO\$\_2\$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO\$*2\$* demand). A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information). If the acclimatory hypothesis is correct, we predict that AAhigh \> AAlow for all species regardless of native habitat or growth environment. Plants adapted to sunny, open habitats will evolve greater stomatal density ratio to take advantage of regular exposure to high light intensity.

Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions (cite). Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments (cites). Plastic changes in leaf anatomy and biochemistry could modulate AA as a byproduct. Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower $g_\mathrm{ias}$; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand. Under the plastic hypothesis, we predict that AAsun \> AAshade for all species and light intensities. Secondarily, AAsun and gsmax,sun should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity. Genotypes adapted to shaded, closed habitats may be plastic, but limits on the width of their reaction norms prevent them from developing traits optimal for conditions they do not regularly experience in nature.

Constitutive hypothesis: In environments that are relatively constant or where environmental change cannot be anticipated by a reliable cue, natural selection will favor constitutive expression of optimal phenotypes. We therefore predict genotypes from more sunny, open habitats will have consistently greater AA under also measurement and growth light intensities. For herbaceous plants, light intensity is largely a function of the tree canopy (cites?). Herbs growing in the open will regularly experience high light intensity; herbs growing under a forest canopy will often experience low light intensity.

CONCEPTUAL FIGURE HERE

@@

## Outtro - I think this goes at the end??
It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat. Reaction norms to light intensity among wild tomatoes indicate that accessions from low light habitats express low \gmaxratio{} when grown in shade, whereas accessions from high light habitats have constitutively higher \gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is driven by greater plasticty in leaf structure in accessions from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis. As a consequence, sun leaves benefit more from amphistomy. This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.
